\bf Model Visualization Techniques for a Social Network Model

### 2017-05-31 paper.Rnw
page 3 edits done

Models for dynamic social networks are extremely important because of how realistic they are. Social networks do not form spontaneously; they evolve over time. Ties are added and removed, and new actors join the network. Modeling the underlying mechanisms that create network changes over time is very complex but also has the potential to uncover hidden truths.

\par One set of models for dynamic social networks are the stochastic actor-oriented models (SAOMs) introduced by \citet{saompaper}. These models are fundamentally different from other social network models because they incorporate network {\emph and} actor statistics into the model, when other models only rely on  the network statistics to determine change. Allowing the actor-level statistics to change the structure of the network leads to a more intuitive model for social network change. In the ``real" world we expect people with common interests to be more likely to form relationships. SAOMs allow us to incorporates this intuition  in the modeling process.

\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}. This software marks a huge contribution to the field of social network analysis.%, and when we began working with it we were eager to understand how it works. HH: we are still interested in this.
However, unlike other models, SAOMs are not very well understood. To get better insight into the \texttt{RSiena} software and the SAOMs themselves, we combine the principals of network visualization with those of model visualization as discussed in \citet{modelvis}. %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
By bringing some light to the underlying methods and structures ``behind the scenes" of SAOMs and RSiena, we  help researchers working with the models better understand and better use these models.

>>>>

Social networks have been studied for decades, beginning with a few foundational works, the most well known of which is the 1967 study, "The Small World Problem" by Stanley Milgram \citep{goldenberg09}. Examples of social networks include collaboration networks between academic researchers, friendship networks in a school or university, and trade networks between nations. In recent years, the study of social networks has grown in popularity due to an increase in the availability and access to social network data. There are many kinds of social networks, but there are not as many statistical models for social network data. Some network models that have been applied to social networks include the exponential random graph model and latent space models. These models, however, are only for single instance networks. If we only care about one network observation, that is not a problem, but if we have many observations of social network over time, these models may not be appropriate. When studying a network over time, which is referred to as a \textit{dynamic network}, we need a model that can take time into account. 
Models for dynamic social networks are extremely important because of how realistic they are. Social networks do not form spontaneously; they evolve over time. Ties are added and removed, and new actors join the network. Modeling the underlying mechanisms that create network changes over time is very complex but also has the potential to uncover hidden truths.

\par In this paper, we concentrate on one type of model for dynamic social networks: the stochastic actor-oriented models (SAOMs), introduced by \citet{saompaper}. These models are fundamentally different from other social network models because they allow us to incorporate network \textit{and} actor statistics, when other models only rely on  the network statistics to determine change. Letting the actor-level statistics change the structure of the network leads to a more practical and relevant approach to model change in a social network: in the ``real" world we expect people with common interests to be more likely to form relationships. SAOMs allow us to incorporate this intuition in the modeling process.

\st{
%\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}. This software marks a huge contribution to the field of social network analysis., and when we began working with it we were eager to understand how it works. HH: we are still interested in this.
Unlike other network models, SAOMs are not very well understood. Traditional modeling steps, like determining the likelihood function, are very complex. Because of this, SAOMs are fit to data using a series of MCMC phases for finding method of moments estimators. Rather than treat SAOM fits and the fitting process as black boxes, we combine the principals of network visualization with those of model visualization as discussed in \citet{modelvis}. %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
By bringing some light to the underlying methods and structures ``behind the scenes" of SAOMs, we help researchers working with the models better understand the implications and analysis of these models.
}
